---
title: "Bioconductor_quiz1"
author: "Shengyuan Wang"
date: "4/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### 1. Use the AnnotationHub package to obtain data on "CpG Islands" in the human genome. Question: How many islands exists on the autosomes?
```{r echo=T, results='hide', message=FALSE, warning=FALSE}
library(IRanges)
library(GenomeInfoDb)
library(GenomicRanges)
library(dbplyr)
library(BiocFileCache)
library(rtracklayer)
library(AnnotationHub)
```

```{r echo=T, message=FALSE, warning=FALSE}
## retrieve the cpg
ah <- AnnotationHub()
ah_human <- subset(ah, species == "Homo sapiens")
ah_human_cpg <- query(ah_human, "CpG Islands")
ah_human_cpg_record <- ah_human_cpg[["AH5086"]]

## extract autosomes
filter <- c(paste("chr", 1:22, sep=""))
split_record <- split(ah_human_cpg_record, seqnames(ah_human_cpg_record))
autosomes <- split_record[filter]

## check the number of autosomes
unlist(autosomes)
```

##### 2. How many CpG Islands exists on chromosome 4.
```{r}
autosomes[4]
```

##### 3. Obtain the data for the H3K4me3 histone modification for the H1 cell line from Epigenomics Roadmap, using AnnotationHub. Subset these regions to only keep regions mapped to the autosomes (chromosomes 1 to 22). How many bases does these regions cover?
```{r echo=T, message=FALSE, warning=FALSE}
## retrieve the record
ah_H3K4me3 <- query(ah, c("H3K4me3", "narrowpeak", "E003"))

## select the narrowpeak
ah_H3K4me3_record <- ah_H3K4me3[["AH29884"]]

## extract autosomes and check the number of regions cover
split_H3K4me3 <- split(ah_H3K4me3_record, seqnames(ah_H3K4me3_record))
H3K4me3_autosomes <- split_H3K4me3[filter]
sum(width(unlist(H3K4me3_autosomes)))
```

##### 4. Obtain the data for the H3K27me3 histone modification for the H1 cell line from Epigenomics Roadmap, using the AnnotationHub package. Subset these regions to only keep regions mapped to the autosomes. In the return data, each region has an associated "signalValue". What is the mean signalValue across all regions on the standard chromosomes?
```{r echo=T, message=FALSE, warning=FALSE}
## retrieve the record
ah_H3K27me3 <- query(ah, c("H3K27me3", "narrowpeak", "E003"))
ah_H3K27me3_record <- ah_H3K27me3[["AH29892"]]

## extract autosomes
split_H3K27me3 <- split(ah_H3K27me3_record, seqnames(ah_H3K27me3_record))
H3K27me3_autosomes <- split_H3K27me3[filter]

## create a subset of extracted autosomes
ah_H3K27me3_autosomes <- subset(ah_H3K27me3_record, seqnames %in% filter)

## mean signalValue
mean_signalValue <- mean(ah_H3K27me3_autosomes$signalValue)
mean_signalValue
```

##### 5. Bivalent regions are bound by both H3K4me3 and H3K27me3. Using the regions we have obtained above, how many bases on the standard chromosomes are bivalently marked?
```{r}
## intersect between two records
bivalent <- intersect(unlist(H3K4me3_autosomes), unlist(H3K27me3_autosomes))
sum(width(bivalent))
```

##### 6. We will examine the extent to which bivalent regions overlap CpG Islands. How big a fraction (expressed as a number between 0 and 1) of the bivalent regions, overlap one or more CpG Islands?
```{r}
# find bivalent regions overlap CpG Islands
cpg_autosomes <- autosomes
cpg_bivalent <- findOverlaps(bivalent, unlist(cpg_autosomes))

# calculate the fraction of the bivalent regions overlap CpG Islands
fraction_bivalent <- length(unique(queryHits(cpg_bivalent)))/length(bivalent)
fraction_bivalent
```

##### 7. How big a fraction (expressed as a number between 0 and 1) of the bases which are part of CpG Islands, are also bivalent marked?
```{r}
cpg_bivalent_intersect <- intersect(bivalent, unlist(cpg_autosomes))

# calculate the fration of the bases intersected between CpG Islands and bivalent
fraction_bivalent_intersect <- sum(width(reduce(cpg_bivalent_intersect)))/sum(width(unlist(cpg_autosomes)))
fraction_bivalent_intersect
```

##### 8. How many bases are bivalently marked within 10kb of CpG Islands?  
Tip: consider using the "resize()"" function.
```{r}
# extract CpG Islands within 10kb
cpg_10kb <- resize(unlist(cpg_autosomes), width = 20000 + width(unlist(cpg_autosomes)), fix = "center")

cpg_10kb_bivalent <- intersect(cpg_10kb, bivalent)
sum(width(cpg_10kb_bivalent))
```

##### 9. How big a fraction (expressed as a number between 0 and 1) of the human genome is contained in a CpG Island?  
Tip 1: the object returned by AnnotationHub contains "seqlengths".  
Tip 2: you may encounter an integer overflow. As described in the session on R Basic Types, you can address this by converting integers to numeric before summing them, "as.numeric()".
```{r}
# calculate human genome size
chr_list <- c(paste("chr", 1:22, sep=""))
genome <- keepSeqlevels(ah_human_cpg_record, chr_list, pruning.mode = "coarse")
genome_size <- sum(as.numeric(seqlengths(genome)))

# calculate the fraction of human genome which contained a CpG Island
cpg_autosomes_size <- sum(as.numeric(width(unlist(cpg_autosomes))))
cpg_autosomes_size / genome_size
```

##### 10. Compute an odds-ratio for the overlap of bivalent marks with CpG islands.
```{r}
## calculate InOut matrix
overlapMat <- matrix(0,, ncol = 2, nrow = 2)
colnames(overlapMat) <- c("in", "out")
rownames(overlapMat) <- c("in", "out")
overlapMat[1,1] <- sum(width(cpg_bivalent_intersect))
overlapMat[1,2] <- sum(width(setdiff(bivalent, unlist(cpg_autosomes))))
overlapMat[2,1] <- sum(width(setdiff(unlist(cpg_autosomes), bivalent)))
overlapMat[2,2] <- genome_size - sum(overlapMat)

## calculate odds-ratio
oddsRatio <- overlapMat[1,1] * overlapMat[2,2] / (overlapMat[2,1] * overlapMat[1,2])
oddsRatio
```
```{r}
sessionInfo()
```

